---
title: "Week 9 Demonstration"
format: html
editor: visual
---

## Set up

```{r}
#| message: FALSE
library(fpp3)
library(tidyverse)
library(slider)
library(gridExtra)
```

## 1. Best linear predictor

Let $X_1,X_2,\ldots,X_n$ be a collection of random variables. For any random variable $Z$, let $Q[Z]$ denote its best linear prediction given $X_1,\ldots,X_n$. Show that $Q$ is a linear operator, i.e. $Q[\alpha X + \beta Y] = \alpha Q[X] + \beta Q[Y]$ for any random variables $X$ and $Y$, and constants $\alpha,\beta$.

## 2. Recursive forecasting

Fit an AR(2) model on the `gtemp_both` dataset filtered to years prior to 1941. Forecast the mean temperature deviation in 1945 using direct forecasting, and via recursive forecasting, showing that they are the same.

```{r}
# Preprocess data
gtemp <- gtemp_both |>
  as_tsibble() |>
  rename(Year = index,
         Temp = value)
```

You may use the following starter code for recursive forecasting

```{r}
# Create dataset with lagged values
gtemp_ar2 <- gtemp_train |>
  # Fill in

# Recursive forecasting
for (i in 1:10) {
  year <- 1940 + i
  idx <- nrow(gtemp_ar2)
  lag1 <- # Fill in
  lag2 <- # Fill in
  pred <- # Fill in
  gtemp_ar2 <- # Fill in
}

gtemp_ar2 |> tail(10)
```

## 3. Nonlinear autoregressive models

Fit a nonlinear AR model using k nearest neighbors on the `gtemp` dataset. Compute its sum of squared residuals and compare it to that of a linear AR model. How about its test error? Use the following starter code.

```{r}
library(caret)

# Create a dataset with columns containing lagged values
gtemp_knn <- gtemp_train |>
  mutate(# Fill in) |>
  filter_index("1854" ~ .)

# Fit KNN model
gtemp_knn_fit <- train(# Fill in, 
  data = gtemp_knn, method = "knn")

# Get training error
ar2_resids <- gtemp_ar2_fit |>
  augment() |>
  pull(.resid)

knn_preds <- predict(gtemp_knn_fit, newdata = gtemp_knn)
true_obs <- gtemp_knn$Temp

print(paste("AR(2) Training MSE:", 
            sprintf("%.4f", mean(ar2_resids[4:length(ar2_resids)] ** 2))))
print(paste("KNN Training MSE:", 
            sprintf("%.4f", mean((true_obs - knn_preds) ** 2))))
```

```{r}
# Recursive forecasting
for (i in 1:10) {
  idx <- nrow(gtemp_knn)
  new_data <- tibble(lag1 = gtemp_knn$Temp[[idx]],
                     lag2 = gtemp_knn$lag1[[idx]],
                     lag3 = gtemp_knn$lag2[[idx]],
                     lag4 = gtemp_knn$lag3[[idx]])
  pred <- predict(gtemp_knn_fit, newdata = new_data)
  new_data <- new_data |> cbind(Year = 1940 + i, Temp = pred)
  gtemp_knn <- gtemp_knn |> add_row(new_data)
}

# Get test error
true_obs <- gtemp |>
  filter_index("1941" ~ "1950") |>
  pull(Temp)

ar2_preds <- gtemp_ar2 |> tail(10) |> pull(Temp)
knn_preds <- gtemp_knn |> tail(10) |> pull(Temp)

print(paste("AR(2) Test MSE:", 
            sprintf("%.4f", mean((true_obs - ar2_preds) ** 2))))
print(paste("KNN Test MSE:", 
            sprintf("%.4f", mean((true_obs - knn_preds) ** 2))))
```

## 4. AR(1)

Recall the calculations for AR(1) in Section 11.3 of the notes.

a.  Show directly that the AR(1) forecast is a best linear predictor.

b.  Show that the estimator we introduced was a method of moments estimator.

c.  What is the conditional maximum likelihood estimate for AR(1)?

## 5. AR(2)

Consider the AR(2) equation $$
X_t = 1.5 X_{t-1} - 0.75 X_{t-2} + W_t.
$$

a.  What is the autoregressive polynomial?

b.  What are its roots?

c.  How can we use the roots to compute the period of its ACF? (optional)

d.  What is the formula for the stationary solution?
