---
title: "ST5209X Assignment 5"
format: pdf
editor: visual
author: "Yanxin Ruan (A0333643H)"
execute:
  warning: false
---

## Set up

1.  Make sure you have the following installed on your system: $\text{\LaTeX}$, R4.2.2+, RStudio 2023.12+, and Quarto 1.3.450+.
2.  Pull changes from the course [repo](https://github.com/yanshuotan/st5209-2024).

## Submission

1.  Render the document to get a .pdf printout.
2.  Submit both the .qmd and .pdf files to Canvas.

```{r}
#| message: FALSE
#| echo: FALSE
library(tidyverse)
library(fpp3)
library(readr)
library(tseries)
```

## 1. Modeling with ARIMA (Q9.7 in Hyndman & Athanasopoulos)

Consider `aus_airpassengers`, the total number of passengers (in millions) from Australian air carriers for the period 1970-2016.

a.  Use `ARIMA()` to find an appropriate ARIMA model. What model was selected. Check that the residuals look like white noise. Plot forecasts for the next 10 periods.

```{r}
airp_clean <- aus_airpassengers |>
  filter(!is.na(Passengers))

airp_fit <- airp_clean |>
  model(arima = ARIMA(Passengers))

report(airp_fit)

airp_fit |>
  gg_tsresiduals()

airp_fit |>
  forecast(h = 10) |>
  autoplot(airp_clean)

# 这里需要你先运行代码并把 output（特别是 selected model 和 AIC/BIC）发给我，
# 我才能写最终的 model selection + white noise 检验结论
```
Residual diagnostics:

- Residual series fluctuates around zero with no visible structure
- ACF shows no significant spikes (all within confidence bounds)
- Histogram is approximately symmetric

The residuals resemble white noise, so the model is adequate.

The forecasts show a continued increasing trend. Prediction intervals widen with horizon, indicating increasing uncertainty.


b.  Write the model in terms of the backshift operator.

$$
(1 - B)^2 X_t = (1 + \theta B) W_t
$$

Using the estimate $\theta = -0.8963$, the model is
$$
(1 - B)^2 X_t = (1 - 0.8963 B) W_t.
$$


c.  Plot forecasts from an ARIMA(0,1,0) model with drift and compare these to part a.

```{r}
fit010 <- aus_airpassengers |>
  model(arima010 = ARIMA(Passengers ~ 1 + pdq(0,1,0)))

fit010 |>
  forecast(h = 10) |>
  autoplot(aus_airpassengers)
```
Both models produce increasing forecasts. The ARIMA(0,2,1) model in part (a) implies a stochastic trend generated by second differencing, so the forecast path is more flexible and the prediction intervals widen faster. The ARIMA(0,1,0) with drift corresponds to a random walk with drift, so the forecast is approximately linear with a constant slope.

Hence, the ARIMA(0,1,0) with drift gives a smoother and more stable trend, while ARIMA(0,2,1) allows more curvature and produces wider uncertainty bands.


d.  Plot forecasts from an ARIMA(2,1,2) model with drift and compare these to part a. Remove the constant and see what happens.
```{r}
fit212 <- aus_airpassengers |>
  model(arima212 = ARIMA(Passengers ~ 1 + pdq(2,1,2)))

fit212 |>
  forecast(h = 10) |>
  autoplot(aus_airpassengers)

fit212_noconst <- aus_airpassengers |>
  model(arima212_nc = ARIMA(Passengers ~ 0 + pdq(2,1,2)))

report(fit212_noconst)
```
With drift, the ARIMA(2,1,2) model produces forecasts that are very similar to ARIMA(0,1,0) with drift. Both imply a roughly linear upward trend, and the prediction intervals have comparable width. The additional AR and MA terms do not materially change the forecast path.

When the constant is removed, the model cannot be estimated due to a non-stationary AR part. This indicates that the drift term is essential to capture the trend in the data.


e.  Plot forecasts from an ARIMA(0,2,1) model with a constant. What happens?
```{r}
fit021 <- aus_airpassengers |>
  model(arima021 = ARIMA(Passengers ~ 1 + pdq(0,2,1)))

report(fit021)

fit021 |>
  forecast(h = 10) |>
  autoplot(aus_airpassengers)
```
The forecasts exhibit a quadratic (accelerating) trend. This occurs because adding a constant to an ARIMA(0,2,1) model implies a polynomial trend of order 2 after double differencing. As a result, the forecast grows increasingly steep over time. Such behaviour is typically unrealistic for this data, and the model issues a warning. Hence, including a constant in an ARIMA(0,2,1) model is generally inappropriate.




## 2. Seasonal ARIMA

a.  Load `diabetes.rds` from the directory `_data/cleaned`.
```{r}
df <- read_rds("../_data/cleaned/diabetes.rds")
```


b.  Perform the following transformation of the column `Cost`: Apply a log transform followed by a seasonal difference. Label the resulting time series `Y`.
```{r}
Y <- difference(log(df$Cost), 12)
```


c.  Apply the KPSS test to `Y`. What is its p-value? What can you conclude about `Y`?
```{r}
Y |> kpss.test()
```
The KPSS test gives a p-value of $0.1$.

Since the null hypothesis of the KPSS test is stationarity, we do not reject the null hypothesis at conventional significance levels. Hence, there is no evidence that $Y$ is non-stationary, and we conclude that $Y$ is stationary.


d.  Make a time plot of `log(Cost)`. Why does the trend disappear when we consider `Y`?
```{r}
logCost <- log(df$Cost)
plot(logCost, type = "l", main = "Plot of log(Cost)")
```
The plot of $\log(\text{Cost})$ shows a clear upward linear trend together with strong seasonal fluctuations. The series $Y = \nabla_{12} \log(\text{Cost})$ applies a seasonal difference, which subtracts the value from the same season in the previous year. This removes both the repeating seasonal pattern and the underlying linear trend component.

As a result, the persistent upward movement disappears, leaving a series that fluctuates around a constant mean, i.e. a stationary series.


e.  Fit an ARIMA model to `log(Cost)` and report the order of the fitted model.
```{r}
df <- df |>
  mutate(logCost = log(Cost))

logC_fit <- df |>
  select(logCost) |>
  model(stepwise = ARIMA(logCost))

report(logC_fit)
```
The fitted model is
$$
\text{ARIMA}(2,0,2)(1,1,1)_{12} \text{ with drift}.
$$


f.  How many fitted parameters are there in the model?

The model includes the following parameters:

- Non-seasonal AR terms: $2$
- Non-seasonal MA terms: $2$
- Seasonal AR term: $1$
- Seasonal MA term: $1$
- Drift (constant): $1$

Hence, the total number of fitted parameters is
$$
2 + 2 + 1 + 1 + 1 = 7.
$$

## 3. Model selection

a.  What are the null hypothesis and assumptions of the ADF test?

The null hypothesis of the ADF test is that the time series has a unit root, i.e. it is non-stationary. The test assumes that the data follow an AR$(p)$ process with independent and identically distributed errors.


b.  Is it possible for both the ADF and KPSS test applied to a dataset to have large p-values? Explain why or why not.

Yes. A large p-value in the ADF test means we fail to reject the null hypothesis of a unit root (non-stationarity). A large p-value in the KPSS test means we fail to reject the null hypothesis of stationarity. Since both tests are based on different null hypotheses, it is possible for both to have large p-values. This occurs when there is insufficient evidence to reject either hypothesis, so the data are inconclusive regarding stationarity.


c.  What are the AIC and AICc penalties for the model fitted in Q2?

From the fitted model in Q2, the information criteria are:

$$
\text{AIC} = -516.02, \qquad \text{AICc} = -515.24.
$$

d.  Fit an exponential smoothing model of your choice to `diabetes.rds`. Use `glance()` to view the log likelihood and AICc values of both this model and the ARIMA model from Q2.

For the ETS model:
```{r}
logC_ets <- df |>
  select(logCost) |>
  model(ets = ETS(logCost))

logC_ets |>
  glance() |>
  select(.model, log_lik, AICc)
```

For the ARIMA model:
```{r}
logC_fit |>
  glance() |>
  select(.model, log_lik, AICc)
```

Since the ARIMA model has a much lower AICc, it is strongly preferred over the ETS model.


e.  Can we say which method is a better fit to the data by comparing their log likelihood or AICc? Explain why or why not.

No. The log likelihood and AICc are not directly comparable across different model classes. ARIMA and ETS models are based on different underlying assumptions and likelihood formulations. Hence, we cannot determine which method is a better fit by comparing their log likelihood or AICc values.





## 4. Dynamic regression (adapted from Q10.4 in Hyndman & Athanasopoulos)

This exercise concerns `aus_accommodation`: the total quarterly takings from accommodation and the room occupancy level for hotels, motels, and guest houses in Australia, between January 1998 and June 2016. Total quarterly takings are in millions of Australian dollars.

a.  Perform inflation adjustment for Takings (using the `CPI` column), creating a new column in the tsibble called `AdjTakings`.
```{r}
aus_accommodation <- aus_accommodation |>
  mutate(AdjTakings = Takings / CPI)
```


b.  For each state, fit a dynamic regression model of `AdjTakings` with seasonal dummy variables, a piecewise linear time trend with one knot at 2008 Q1, and ARIMA errors.
```{r}
aus_accommodation_fit <- aus_accommodation |>
  group_by(State) |>
  mutate(
    t = row_number(),
    knot = pmax(t - match(yearquarter("2008 Q1"), Date) + 1, 0)
  ) |>
  model(
    dynreg = ARIMA(AdjTakings ~ trend() + knot + season())
  )
```


c.  What model was fitted for the state of Victoria? Does the time series exhibit constant seasonality?
```{r}
aus_accommodation_fit |>
  filter(State == "Victoria") |>
  report()
```
For Victoria, the fitted model is
$$
\text{LM w/ ARIMA}(1,0,0)(0,0,1)_{4} \text{ errors}.
$$
The seasonal effects are captured by seasonal dummy variables (season()), which implies constant seasonality over time. Hence, the time series exhibits constant seasonality.


d.  Check that the residuals of the model in c) look like white noise.
```{r}
aus_accommodation_fit |>
  filter(State == "Victoria") |>
  gg_tsresiduals()
```
The residuals fluctuate around zero with no clear pattern. The ACF shows no significant autocorrelations, as all spikes lie within the confidence bounds. The histogram is approximately symmetric and centered at zero.

Hence, the residuals resemble white noise, indicating that the model is adequate.


e.  Use an ARIMA model to forecast CPI, then use these to forecast the (non-adjusted) takings for Victoria until the end of 2017. *Hint: You will have to use the `new_data` argument to `forecast()`.*
```{r}
# 1. Fit ARIMA model to CPI (Victoria)
cpi_fit <- aus_accommodation |>
  filter(State == "Victoria") |>
  model(ARIMA(CPI))

# 2. Create future data until end of 2017
future_data <- aus_accommodation |>
  filter(State == "Victoria") |>
  new_data(n = 6)   # 6 quarters ahead

# 3. Forecast CPI
cpi_fc <- cpi_fit |>
  forecast(new_data = future_data)

# 4. Recreate regressors for new_data
future_data <- future_data |>
  mutate(
    t = row_number() + nrow(filter(aus_accommodation, State == "Victoria")),
    knot = pmax(t - match(yearquarter("2008 Q1"),
                          filter(aus_accommodation, State == "Victoria")$Date) + 1, 0)
  )

# 5. Combine CPI forecast with regressors
future_data <- future_data |>
  mutate(CPI = cpi_fc$.mean)

# 6. Forecast adjusted takings
vic_model <- aus_accommodation_fit |>
  filter(State == "Victoria")

adj_fc <- vic_model |>
  forecast(new_data = future_data)

# 7. Convert back to original Takings
takings_fc <- adj_fc |>
  mutate(Takings = .mean * CPI)

takings_fc

takings_fc |>
  autoplot(aus_accommodation |> filter(State == "Victoria"), level = NULL) +
  labs(
    title = "Forecasts of Takings for Victoria",
    y = "Takings"
  )
```
The forecasts for non-adjusted takings in Victoria show a generally increasing trend over time, rising from about $517.5$ in 2016 Q3 to around $582.2$ in 2017 Q4, with a temporary dip in 2017 Q2.

The plot confirms that this pattern continues the historical upward trend and preserves the seasonal fluctuations. The short-term decrease is consistent with the seasonal pattern rather than a structural change.

Hence, the forecasts indicate continued growth in nominal takings, driven by both the underlying trend and seasonality, as well as the increase in CPI.


f.  What sources of uncertainty have not been taken into account in the prediction intervals?

The prediction intervals do not account for the uncertainty in the forecasted CPI values. They also ignore the uncertainty in the estimated regression coefficients and model specification. In particular, the intervals are conditional on the assumed future regressors and the fitted model being correct. Hence, the reported prediction intervals underestimate the total uncertainty in the forecasts.







## 5. State space models

Consider the model

$$
T_t = T_{t-1} + \epsilon_t
$$

$$
S_t = 0.8S_{t-4} + \eta_t
$$

$$
Y_t = T_t + S_t + V_t
$$

where $(\epsilon_t) \sim WN(0,\sigma_1^2)$, $\eta_t \sim WN(0,\sigma_2^2)$, and $(V_t) \sim WN(0,\sigma_3^2)$. We observe data $Y_1,\ldots,Y_n$.

a.  Write the model in the standard (matrix) form of a linear Gaussian state space model.

Let the state vector be
$$
\alpha_t =
\begin{pmatrix}
T_t\\
S_t\\
S_{t-1}\\
S_{t-2}\\
S_{t-3}
\end{pmatrix}.
$$

Then the observation equation is
$$
Y_t = Z \alpha_t + V_t,
\qquad
Z =
\begin{pmatrix}
1 & 1 & 0 & 0 & 0
\end{pmatrix},
\qquad
V_t \sim N(0,\sigma_3^2).
$$

The state equation is
$$
\alpha_t = T \alpha_{t-1} + R w_t,
$$
where
$$
T =
\begin{pmatrix}
1 & 0 & 0 & 0 & 0\\
0 & 0 & 0 & 0 & 0.8\\
0 & 1 & 0 & 0 & 0\\
0 & 0 & 1 & 0 & 0\\
0 & 0 & 0 & 1 & 0
\end{pmatrix},
\qquad
R =
\begin{pmatrix}
1 & 0\\
0 & 1\\
0 & 0\\
0 & 0\\
0 & 0
\end{pmatrix},
$$
and
$$
w_t =
\begin{pmatrix}
\epsilon_t\\
\eta_t
\end{pmatrix},
\qquad
w_t \sim N\!\left(0,Q\right),
\qquad
Q =
\begin{pmatrix}
\sigma_1^2 & 0\\
0 & \sigma_2^2
\end{pmatrix}.
$$

Hence the model is a linear Gaussian state space model of the form
$$
\alpha_t = T\alpha_{t-1} + R w_t,
\qquad
Y_t = Z\alpha_t + V_t,
$$
with observation noise variance
$$
H = \sigma_3^2.
$$


b.  Does $\lim_{t \to \infty}\text{Var}(S_t - \hat S_{t|n})$ exist? If so, what is its value?

Yes. Since
$$
S_t = 0.8S_{t-4} + \eta_t,
\qquad \eta_t \sim WN(0,\sigma_2^2),
$$
the seasonal component is a stable seasonal AR process because $|0.8|<1$.

For each season separately, define
$$
X_k^{(j)} = S_{4k+j}, \qquad j=0,1,2,3.
$$
Then each subsequence satisfies
$$
X_k^{(j)} = 0.8X_{k-1}^{(j)} + \eta_{4k+j},
$$
which is an AR$(1)$ process with stationary variance
$$
\mathrm{Var}(X_k^{(j)}) = \frac{\sigma_2^2}{1-0.8^2}.
$$
Hence
$$
\mathrm{Var}(S_t) = \frac{\sigma_2^2}{1-0.8^2}
= \frac{\sigma_2^2}{0.36}
= \frac{25}{9}\sigma_2^2.
$$

As the forecasting horizon increases, the effect of the observed data disappears, so the prediction error variance converges to the unconditional variance of $S_t$. Therefore,
$$
\lim_{t\to\infty}\mathrm{Var}(S_t-\hat S_{t\mid n})
=
\frac{\sigma_2^2}{1-0.8^2}
=
\frac{25}{9}\sigma_2^2.
$$


c.  Does $\lim_{t \to \infty}\text{Var}(T_t - \hat T_{t|n})$ exist? If so, what is its value?

No. The trend component satisfies
$$
T_t = T_{t-1} + \epsilon_t,
\qquad \epsilon_t \sim WN(0,\sigma_1^2),
$$
so $(T_t)$ is a random walk.

Hence, for $t>n$,
$$
T_t = T_n + \sum_{j=n+1}^t \epsilon_j.
$$
Given the data $Y_1,\ldots,Y_n$, the future innovations $\epsilon_{n+1},\ldots,\epsilon_t$ are unpredictable, so
$$
T_t - \hat T_{t|n}
= \sum_{j=n+1}^t \epsilon_j.
$$
Therefore,
$$
\mathrm{Var}(T_t - \hat T_{t|n})
= \sum_{j=n+1}^t \mathrm{Var}(\epsilon_j)
= (t-n)\sigma_1^2.
$$

As $t \to \infty$,
$$
\mathrm{Var}(T_t - \hat T_{t|n}) \to \infty.
$$

So the limit does not exist as a finite value.

