---
title: "ST5209X Assignment 4"
format: pdf
editor: visual
author: "Yanxin Ruan (A0333643H)"
execute:
  warning: false
---

## Set up

1.  Make sure you have the following installed on your system: $\text{\LaTeX}$, R4.2.2+, RStudio 2023.12+, and Quarto 1.3.450+.
2.  Pull changes from the course [repo](https://github.com/yanshuotan/st5209-2024).

## Submission

1.  Render the document to get a .pdf printout.
2.  Submit both the .qmd and .pdf files to Canvas.

```{r}
#| message: FALSE
#| echo: FALSE
library(tidyverse)
library(fpp3)
library(tsibble)
library(fable)
library(feasts)
library(dplyr)
library(tidyr)
library(broom)
```

## 1. AR polynomial

Consider the AR(2) model 
$$
X_t = 4 + 0.5 X_{t-1} - 0.25 X_{t-2} + W_t.
$$ {#eq-ar2}

a.  What is the autoregressive polynomial?

Since the standard AR(2) form is
$$
X_t = c + \phi_1 X_{t-1} + \phi_2 X_{t-2} + W_t.
$$

By direct comparison,
$$
\phi_1 = 0.5, \qquad \phi_2 = -0.25.
$$

For an AR(2) model, the autoregressive polynomial is defined by
$$
\phi(z) = 1 - \phi_1 z - \phi_2 z^2.
$$

Substitute the values of $\phi_1$ and $\phi_2$:
$$
\phi(z) = 1 - 0.5z - (-0.25)z^2.
$$

Now simplify the sign:
$$
\phi(z) = 1 - 0.5z + 0.25z^2.
$$

Therefore, the autoregressive polynomial is
$$
\phi(z) = 1 - 0.5z + 0.25z^2.
$$


b.  What are its roots?

We solve
$$
\begin{aligned}
\phi(z) = 1 - 0.5z + 0.25z^2 &= 0. \\
z^2 - 2z + 4 &= 0.
\end{aligned}
$$

By applying the quadratic formula, we get
$$
z = \frac{-(-2) \pm \sqrt{(-2)^2 - 4\times 1 \times 4}}{2 \times 1}.
$$

Therefore, the two roots are
$$
z_1 = 1 + \sqrt{3}\,i, \qquad z_2 = 1 - \sqrt{3}\,i.
$$


c.  Is this model causal? Why?

An AR model is causal if all roots of the autoregressive polynomial satisfy
$$
|z| > 1.
$$

We compute the modulus of each root as follows.

For $z_1$,
$$
|z_1| = |1 + \sqrt{3}\,i|
= \sqrt{1^2 + (\sqrt{3})^2}
= 2 > 1.
$$

Similarly, for $z_2$,
$$
|z_2| = |1 - \sqrt{3}\,i| = \sqrt{1^2 + (-\sqrt{3})^2}
= 2 > 1.
$$

Thus, both roots lie outside the unit circle. Therefore, the model is **causal**.


d.  Given the representation $$
    X_t = \mu + \sum_{j=0}^\infty \psi_j W_{t-j},
    $$ solve for $\psi_0, \psi_1, \psi_2, \psi_3$. *Hint: Follow the strategy discussed in `linear_process_coefs.qmd`.*

Take expectation on both sides of the equation of $X_t$. Since $E(W_t)=0$, and for a stationary causal process
$$
E(X_t) = E(X_{t-1}) = E(X_{t-2}) = \mu,
$$
we get
$$
\begin{aligned}
\mu &= 4 + 0.5\mu - 0.25\mu + 0 \\
0.75\mu &= 4 \\
\mu &= \frac{16}{3}
\end{aligned}
$$

For a causal AR process,
$$
X_t - \mu = \sum_{j=0}^{\infty} \psi_j W_{t-j},
$$
where the coefficients satisfy
$$
\phi(z)\left(\sum_{j=0}^{\infty} \psi_j z^j\right) = 1.
$$

By substituting $\phi(z)$ we derived in part (a), we get
$$
(1 - 0.5z + 0.25z^2)\left(\sum_{j=0}^{\infty} \psi_j z^j\right) = 1.
$$

Now expand the product carefully:
$$
\left(1 - 0.5z + 0.25z^2\right)\left(\psi_0 + \psi_1 z + \psi_2 z^2 + \psi_3 z^3 + \cdots\right) = 1.
$$

Distribute each term:

From $1$,
$$
1 \cdot (\psi_0 + \psi_1 z + \psi_2 z^2 + \psi_3 z^3 + \cdots)
= \psi_0 + \psi_1 z + \psi_2 z^2 + \psi_3 z^3 + \cdots
$$

From $-0.5z$,
$$
-0.5z \cdot (\psi_0 + \psi_1 z + \psi_2 z^2 + \psi_3 z^3 + \cdots)
= -0.5\psi_0 z - 0.5\psi_1 z^2 - 0.5\psi_2 z^3 - 0.5\psi_3 z^4 - \cdots
$$

From $0.25z^2$,
$$
0.25z^2 \cdot (\psi_0 + \psi_1 z + \psi_2 z^2 + \psi_3 z^3 + \cdots)
= 0.25\psi_0 z^2 + 0.25\psi_1 z^3 + 0.25\psi_2 z^4 + 0.25\psi_3 z^5 + \cdots
$$

Now collect coefficients of equal powers of $z$.

$$
\begin{aligned}
&\text{The constant term is } &\quad &\psi_0 \\
&\text{The coefficient of }z \text{ is } &\quad &\psi_1 - 0.5\psi_0 \\
&\text{The coefficient of }z^2 \text{ is } &\quad &\psi_2 - 0.5\psi_1 + 0.25\psi_0 \\
&\text{The coefficient of }z^3 \text{ is } &\quad &\psi_3 - 0.5\psi_2 + 0.25\psi_1
\end{aligned}
$$

Therefore,
$$
(1 - 0.5z + 0.25z^2)\sum_{j=0}^{\infty}\psi_j z^j
=
\psi_0
+ (\psi_1 - 0.5\psi_0)z
+ (\psi_2 - 0.5\psi_1 + 0.25\psi_0)z^2
+ (\psi_3 - 0.5\psi_2 + 0.25\psi_1)z^3
+ \cdots
$$

But this product must equal $1$, and
$$
1 = 1 + 0z + 0z^2 + 0z^3 + \cdots
$$

So the coefficients must match term by term.

- For the constant term:
$$
\psi_0 = 1.
$$

- For the coefficient of $z$:
$$
\psi_1 - 0.5\psi_0 = 0.
$$

Substitute $\psi_0=1$:
$$
\begin{aligned}
\psi_1 - 0.5(1) &= 0 \\
\psi_1 &= 0.5
\end{aligned}
$$

- For the coefficient of $z^2$:
$$
\psi_2 - 0.5\psi_1 + 0.25\psi_0 = 0.
$$

Substitute $\psi_1=0.5$ and $\psi_0=1$:
$$
\begin{aligned}
\psi_2 - 0.5(0.5) + 0.25(1) &= 0 \\
\psi_2 &= 0
\end{aligned}
$$

- For the coefficient of $z^3$:
$$
\psi_3 - 0.5\psi_2 + 0.25\psi_1 = 0.
$$

Substitute $\psi_2=0$ and $\psi_1=0.5$:
$$
\begin{aligned}
\psi_3 - 0.5(0) + 0.25(0.5) &= 0 \\
\psi_3 &= -0.125 = -\frac{1}{8}
\end{aligned}
$$

Therefore,
$$
\mu = \frac{16}{3},
$$
and the first four coefficients are
$$
\psi_0 = 1, \qquad
\psi_1 = 0.5, \qquad
\psi_2 = 0, \qquad
\psi_3 = -\frac{1}{8}.
$$






## 2. Likelihood

Consider the AR(1) model $$
X_t = 1-0.6X_{t-1} + W_t,
$$ where $W_t \sim WN(0, 0.25)$ is Gaussian white noise. We are given observations $x_1 = 0.2, x_2 = -0.3, x_3 = 0.4$.

a.  What is the mean of $X_t$?

Let
$$
\mu = E(X_t).
$$

Taking expectation on both sides and using $E(W_t)=0$ gives
$$
\mu = 1 - 0.6\mu.
$$

Therefore, the mean of $X_t$ is
$$
\mu = 0.625.
$$


b.  Write the conditional likelihood of this model, when conditioning on the value of $x_1$.

Conditioning on $x_1$, the conditional likelihood is built from the conditional densities
$$
f(x_2,x_3\mid x_1)=f(x_2\mid x_1)f(x_3\mid x_2).
$$

From the model,
$$
X_t \mid X_{t-1}=x_{t-1} \sim N(1-0.6x_{t-1},\,0.25).
$$

Therefore,
$$
X_2 \mid X_1=x_1 \sim N(1-0.6x_1,\,0.25),
\qquad
X_3 \mid X_2=x_2 \sim N(1-0.6x_2,\,0.25).
$$

So the conditional likelihood is
$$
L_c
= f(x_2\mid x_1)f(x_3\mid x_2)
= \prod_{t=2}^3 \frac{1}{\sqrt{2\pi(0.25)}}
\exp\!\left(
-\frac{(x_t-(1-0.6x_{t-1}))^2}{2(0.25)}
\right).
$$

Substituting $x_1=0.2$, $x_2=-0.3$, $x_3=0.4$,
$$
\begin{aligned}
L_c
&=
\frac{1}{\sqrt{2\pi(0.25)}}\exp\!\left(
-\frac{(-0.3-(1-0.6\cdot 0.2))^2}{2(0.25)}
\right)
\cdot
\frac{1}{\sqrt{2\pi(0.25)}}\exp\!\left(
-\frac{(0.4-(1-0.6(-0.3)))^2}{2(0.25)}
\right) \\
&=
\frac{1}{2\pi(0.25)}
\exp\!\left(
-\frac{1}{2(0.25)}
\left[
(-0.3-0.88)^2 + (0.4-1.18)^2
\right]
\right) \\
&\approx 0.0116 \quad \text{ (to 4 d.p.)}
\end{aligned}
$$







## 3. Reversibility

a.  Using `set.seed(5209)` to fix a random seed, create a sample trajectory of length $n = 500$ from the AR(2) model from Problem 1 using `arima.sim()`.

Simulate from the AR(2) model
$$
X_t = 4 + 0.5X_{t-1} - 0.25X_{t-2} + W_t.
$$

In `arima.sim()`, the argument `mean = 4` directly incorporates the intercept, so no manual centering is needed. The coefficients are specified as
$$
\phi_1 = 0.5, \qquad \phi_2 = -0.25.
$$
```{r}
set.seed(5209)

ar2 <- arima.sim(
  model = list(ar = c(0.5, -0.25)),
  n = 500,
  mean = 4
)

ar2_df <- tibble(
  time = 1:500,
  value = as.numeric(ar2)
)

ggplot(ar2_df, aes(x = time, y = value)) +
  geom_line() +
  labs(x = "time", y = "value")
```


b.  Reverse the time index of the vector you obtain using `rev()`.

Reverse the time index using `rev()`. This produces a series that runs backward in time.
```{r}
ar2_rev <- rev(ar2)

ar2_rev_df <- tibble(
  time = 1:500,
  value = as.numeric(ar2_rev)
)

ggplot(ar2_rev_df, aes(x = time, y = value)) +
  geom_line() +
  labs(x = "time", y = "value")
```


c.  Fit an AR(2) model to the reversed time series using `fable`. Hint: You may use the code snippet `model(AR(X ~ order(2)))`.

```{r}
rev_ts <- ar2_rev_df %>%
  as_tsibble(index = time)

fit_rev <- rev_ts %>%
  model(AR(value ~ order(2)))
```


d.  Inspect the model parameters using `tidy()`. Why are they similar to the those in @eq-ar2?

Inspect the estimated parameters using `tidy()`. The estimated coefficients are expected to be close to
$$
\phi_1 = 0.5, \qquad \phi_2 = -0.25,
$$
because a stationary Gaussian AR(2) process is reversible in distribution. Therefore, reversing time does not change the second-order dependence structure, and estimation recovers similar parameters up to sampling variation.
```{r}
tidy(fit_rev)
```

e.  Make a forecast with $h = 10$. What does this correspond to in terms of the original time series?

```{r}
fc_rev <- fit_rev %>%
  forecast(h = 10)

fc_rev
autoplot(fc_rev, rev_ts)
```

The forecast indices are 501--510 for the reversed series. These correspond to $X_0, X_{-1}, \ldots, X_{-9}$ in the original series, i.e. backcasts before $X_1$.







## 4. Yule-Walker

a.  Write the Yule-Walker equations for the AR(2) model from Problem 1.

b.  Arrange the equations in the following matrix form (i.e. fill in the missing entries): $$
    \left[\begin{matrix}     ? & ? & ?  \\ ? & ? & ? \\
    ? & ? & ?
    \end{matrix}\right] \left[\begin{matrix}     \gamma(0) \\ \gamma(1) \\ \gamma(2)
    \end{matrix}\right] = \left[\begin{matrix}     \sigma^2 \\ 0 \\ 0
    \end{matrix}\right].
    $$

c.  Solve the system from part b) for $\gamma(0), \gamma(1), \gamma(2)$ numerically using `solve()`.

d.  Given $$
    \Gamma_2 = \left[\begin{matrix}     \gamma(0) & \gamma(1) \\ \gamma(1) & \gamma(0)
    \end{matrix}\right],
    $$ what is the top left entry of $\Gamma_2^{-1}$ in terms of $\gamma(0)$ and $\gamma(1)$?

e.  Write a 95% confidence interval for $\phi_1$ using your answers for d) and 3d).

## 5. Recursive forecasting for ARMA(1, 1)

Consider the $ARMA(1,1)$ model $$
X_t - 0.5X_{t-1} = W_t + 0.5W_{t-1}.
$$ In this question, we will investigate recursive forecasting. The following code snippet generates a sequence of length $n=50$ drawn from the above model.

```{r}
set.seed(5209)
n <- 50
wn <- rnorm(n)
xt <- arima.sim(model = list(ar = 0.5, ma = 0.5), innov = wn, n = n)
```

a.  Fill in the following code snippet using equation (12.14) to generate a sequence `wn_hat`.

```{r}
wn_hat <- rep(0, n)
wn_hat[[1]] <- xt[[1]]
for (i in 2:n) {
  # FILL IN
}
```

b.  Make a time plot of the log absolute difference between `wn` and `wn_hat`.

c.  What consequence does this have for truncated forecasts?

d.  Compute the truncated forecast for $X_{53}$.

## 6. ACF, PACF, and BLPs

Let $(X_t)$ be a mean zero stationary process with the following autocovariance values:

$$
\gamma_X(0) = 2, \gamma_X(1) = 1.4, \gamma_X(2) = 0.6, \gamma_X(3) = 0.4, \gamma_X(4) = 0.2.
$$

a.  Can $(X_t)$ be an MA(2) process? Explain why or why not.

b.  Can $(X_t)$ be an AR(1) process? Explain why or why not.

c.  What is the best linear predictor $\hat X_4$ for $X_4$ given only $X_3 = 2$?

d.  Using the notation in part c), what is the variance of $X_4 - \hat X_4$?

e.  What is the best linear predictor $\hat X_4$ for $X_4$ given only $X_2 = 2$?

f.  Using the notation in part e), what is the variance of $X_4 - \hat X_4$?

g.  Let $\alpha_X$ denote the partial autocorrelation function of $(X_t)$. What is $\alpha_X(1)$?

h.  What is $\alpha_X(3)$?
